The \project{unRAVE} catalog

### 2016-09-08 article/unrave.tex
calibration comparison

We calculated all possible pair-wise differences between the labels we
derived from multiple visits.  If \rave\ observed a star $H$ times, there 
are $H!/2(H-2)!$ possible $a$-to-$b$ pair-wise combinations where we can calculate 
the difference between the derived label (for example, $\logg$) over the
quadrature sum of their formal uncertainties $(\logg_a - \logg_b)/\sqrt{{\sigma_{\logg,a}}^2 + {\sigma_{\logg,b}}^2}$.  If our derived labels and formal errors were 
correct, the distribution of these pair-wise comparisons would be
well-represented by a Gaussian distribution with zero mean and variance
of unity.  However, our formal errors are likely to be under-estimated,
and therefore we introduce a systematic error floor for each label, which
is added in quadrature to every observation such that (for example, $\logg$),
\begin{eqnarray}
	\eta_{\logg} & = & \frac{\logg_a - \logg_b}{\sqrt{(\sigma_{\logg,a} + \sigma_{\logg,floor})^2 + (\sigma_{\logg,b} + \sigma_{\logg,floor})^2}} \quad .
\end{eqnarray}

@@

We further used the repeat observations in \rave\ to build intuition for
the label precision that was achievable as a function of S/N ratio.  
Specifically we stacked all spectra for a given star by summing the 
fluxes weighted by the inverse variances, then treated the stacked spectra
as a normal survey star.  We inferred labels for the stacked spectra 
using all three models, and derived a joint estimate as per Section 
\ref{sec:joining-the-models}.  The labels we inferred from the stacked 
spectrum then served as a basis of comparison for the labels we derived 
from the individual visit spectra of the same star, which are of lower 
S/N ratios.

In Figure \ref{fig:repeat-visits} we show the RMS difference
in labels between the stacked spectrum and a single visit, binned by the
S/N ratio of the individual visit spectrum.  Here we only show stars
where the \emph{stacked} spectrum had S/N $>100$~pixel$^{-1}$ to ensure
that our baseline comparisons were in a region where we are dominated
by systematic uncertainties.  The precision in all labels tends to
flatten out past S/N $> 40$~pixel$^{-1}$, and the precision at high S/N
ratios is comparable to the minimum error floors we adopted in Section
\ref{sec:repeat-observations}.  The median S/N of \rave\ spectra is
50~pixel$^{-1}$, at which point our abundance precision is $\approx0.1$~dex,
varying a few tenths of a dex between elements.

@@

We searched the literature for studies that overlap with \rave, and which base
their analysis on high-resolution, high S/N spectra.  We found four notable
studies with a sufficient level of overlap: the disk studies by \citet{Reddy_2003,Reddy_2006}
and \citet{Bensby_2014}, as well as the \citet{Valenti_Fischer_2005} work on exoplanet
host star candidates.  These studies 
perform a careful (manual; expert) analysis using extremely high-resolution, high S/N spectra, and make use of \project{Hipparcos}
parallaxes where possible.  Most of the stars in these samples are main-sequence
or sub-giant stars.  Therefore, these works constitute an excellent comparison
to evaluate the accuracy of our results on the main-sequence and sub-giant branch.

We compare the stellar parameters ($\teff$, $\logg$, [Fe/H]) of stars that
overlap these literature studies in Figure~\ref{fig:gold-standard-comparison}.
We only include stars with $\chi_r^2 < 3$ and $S/N > 10$~pixel$^{-1}$, although the latter
cut removed only a few stars because the average S/N in the \rave\ spectra for
these stars is relatively high ($\gtrsim{}50$~pixel$^{-1}$). When considering the relative 
information content available in \rave\ (945~pixels in the near infrared with
$\mathcal{R} \approx 7{,}500$) compared to these literature studies that use
\project{Hipparcos} parallaxes where possible, and base their inferences on
spectra with resolving power $\mathcal{R}$ between 40,000 to 110,000, and
S/N ratios exceeding 150, we consider the agreement to be very satisfactory.
Indeed, given the metallicity precision available in the \unrave\ catalog, 
these results will likely be useful for future studies based on exoplanet
host star properties (e.g., \project{TESS}). At present, however, there are 
just $\approx$30 stars in \rave\ that overlap with the compilations of 
exoplanet host star properties listed at \texttt{exoplanets.org} and 
\texttt{exoplanets.eu}.

\stub{Reference Figure \ref{fig:gold-standard-hrd}}

@@

There are 142 stars that overlap between \rave\ and the fourth internal
data release of the \ges\ survey. These are a mix of main-sequence, 
sub-giant and red giant branch stars.  About half (67) of the sample 
were acquired with the \acronym{UVES} instrument, the other with the \acronym{GIRAFFE} spectrograph, and
the S/N of the \ges\ spectra peaks at $\approx140$~pixel$^{-1}$.  Despite 
most of these stars having
relatively low S/N ratios in \rave\ ($\approx 25$~pixel$^{-1}$), there is good 
agreement in with \ges\ and the \unrave\ stellar parameters (Figure~\ref{fig:ges-stellar-parameters}).  
The RMS in effective temperature, surface gravity and metallicity is
233~K, 0.37~dex, and 0.17~dex, respectively.

Based on this comparison, we find no evidence for a systematic offset 
between stars on the main-sequence and those on the giant branch. Specifically,
we do not find any dwarf stars that are misclassified as sub-giants, as we
found for some metal-poor stars in Section~\ref{sec:validation-gold-standards}.
In addition we find no evidence for a systematic offset in metallicity (or
individual abundances) between giant stars and main-sequence stars.  This is
a crucial observation, as the metallicities for stars in our main-sequence
training set have a principally different source to those on the giant branch.
We cannot make these same inferences based on other surveys, like \apogee,
because (1) \apogee\ stars formed part of the training set, and (2) they do
not include main-sequence stars.  Even if we found good agreement between
\epic\ and \apogee\ metallicities, this would not be informative, because
\apogee\ is the source of metallicity for many stars on the giant branch in
the \epic\ sample. Therefore, although this is a qualitative comparison only,
it is reassuring that there is no obvious systematic difference between the
metallicities of main-sequence and giant branch stars.

The metallicity agreement between this work and \ges\ extends down to low
metallicity, near $[{\rm Fe/H}] \approx -1.5$.  The scatter increases for
the four stars in the overlap sample with [Fe/H] $< -1$, however these
spectra have low S/N ratios, which is reflected in the larger uncertainties
reported for these metallicities. 
% TODO: Check if that is still the case after we examine errors from repeat
% 		obs.

The fourth internal data release of the \ges\ includes detailed chemical
abundances of up to 45~species ($\approx32$ elements at different ionization
stages).  This provides us with an independent validation for our detailed
abundances on the giant branch.  These comparisons are shown for six elemental
abundances in Figure~\ref{fig:ges-abundances}, where markers are colored by the
S/N of the \rave\ spectrum.  The number of stars available in each abundance
comparison varies due to what is available in the \ges\ data release, which
is itself a function of the instrument used, the spectral type, and other
factors.  The absolute bias for individual elements varies from as low as
0.01~dex ([Al/H]) to as high as 0.22~dex ([Si/H]), where we over-estimate 
[Si/H] abundances relative to the \ges\ survey.  The large bias in [Si/H]
is likely a consequence of an offset between [Si/H] abundances in \ges\ and
\apogee, the source of our training set for giant star abundances.  
The RMS deviation in each
label is small for stars with [X/H] $> -0.5$, before increasing at lower
metallicities.  If we consider all stars, the smallest abundance RMS we see
with respect to \ges\ is 0.18~dex for [Ca/H].  The increasing RMS at low 
metallicity is likely a consequence of multiple factors, namely: inaccurate
abundance labels for metal-poor stars (Section~\ref{sec:the-training-set});
only weak, blended lines being available in \rave, which cease to be visible
in hot and/or metal-poor stars; and to a lesser extent, low S/N ratios for 
those particular stars being compared.  Unfortunately not all of these
factors are represented by the quoted errors in each label.  For these reasons, 
although it affects only a small number of stars, we recommend caution when 
using individual abundances for very metal-poor giant stars in our sample.

\subsubsection{Comparison with the \rave\ DR4 calibration sample}

\stub{Reference Figure\ref{fig:kordopatis-calibration}}

>>>>

We calculated all possible pair-wise differences between the labels we
derived from multiple visits.  If \rave\ observed a star $H$ times, there 
are $H!/2(H-2)!$ possible $a$-to-$b$ pair-wise combinations where we can calculate 
the difference between the derived label (for example, $\logg$) over the
quadrature sum of their formal uncertainties $(\logg_a - \logg_b)/\sqrt{{\sigma_{\logg,a}}^2 + {\sigma_{\logg,b}}^2}$.  If our derived labels were unbiased and our formal
errors were correct, the distribution of these pair-wise comparisons would be
well-represented by a Gaussian distribution with zero mean and variance
of unity.  However, our formal errors are likely to be under-estimated,
and therefore we introduce a systematic error floor for each label, which
is added in quadrature to every observation such that (for example, $\logg$),
\begin{eqnarray}
	\eta_{\logg} & = & \frac{\logg_a - \logg_b}{\sqrt{{\sigma_{\logg,a}}^2 + {\sigma_{\logg,b}}^2 + 2{\sigma_{\logg,floor}}^2}} \quad .
\end{eqnarray}

@@

We further used the repeat observations in \rave\ to build intuition for
the label precision that was achievable as a function of S/N ratio.  
Specifically we stacked all spectra for a given star by summing the 
fluxes weighted by the inverse variances, then treated the stacked spectra
as normal survey stars.  We inferred labels for the stacked spectra 
using all three models, and derived a joint estimate as per Section 
\ref{sec:joining-the-models}.  The labels we inferred from each stacked 
spectrum then served as a basis of comparison for the labels we derived 
from the individual visit spectra of the same star, which are of lower 
S/N ratios.

In Figure \ref{fig:repeat-visits} we show the RMS difference
in labels between the stacked spectra and single visit, binned by the
S/N ratio of the individual visit spectrum.  Here we only show stars
where the \emph{stacked} spectrum had S/N $>100$~pixel$^{-1}$ to ensure
that our baseline comparisons were in a region where we are dominated
by systematic uncertainties.  The precision in all labels tends to
flatten out past S/N $> 40$~pixel$^{-1}$, and the precision at high S/N
ratios is comparable to the minimum error floors we adopted in Section
\ref{sec:repeat-observations}.  The median S/N of \rave\ spectra is
50~pixel$^{-1}$, at which point our abundance precision is $\approx0.1$~dex,
varying a few tenths of a dex between elements.

@@

We searched the literature for studies that overlap with \rave, and which base
their analysis on high-resolution, high S/N spectra.  We found four notable
studies with a sufficient level of overlap: the disk studies by \citet{Reddy_2003,
Reddy_2006} and \citet{Bensby_2014}, as well as the \citet{Valenti_Fischer_2005} 
work on exoplanet host star candidates.  These studies perform a careful (manual;
expert) analysis using extremely high-resolution, high S/N spectra, and make use 
of \project{Hipparcos} parallaxes where possible.  Most of the stars in these 
samples are main-sequence or sub-giant stars.  Therefore, these works constitute 
an excellent comparison to evaluate the accuracy of our results on the 
main-sequence and sub-giant branch.

In Figure \ref{fig:gold-standard-hrd} we show Hertzsprung-Russell diagrams 
for the \rave\ stars that overlap with these studies.  We only include stars 
with $\chi_r^2 < 3$ and $S/N > 10$~pixel$^{-1}$, although the latter cut 
removed only a few stars because the average S/N in the \rave\ spectra for
these stars is relatively high ($\gtrsim{}50$~pixel$^{-1}$).  The literature 
data points in Figure \ref{fig:gold-standard-hrd} are linked to our derived
labels for the same stars, illustrating good qualitative agreement across 
the turn-off and sub-giant branch in all studies.  If we treat all three 
studies as a single point of comparison, the bias between our work and these 
studies is $-89$~K in $\teff$, just $-0.06$~dex in $\logg$, and $-0.03$~dex 
in [Fe/H] (see Figure~\ref{fig:gold-standard-comparison}).  The RMS deviation
in labels is 237~K, 0.30~dex, and 0.15~dex, respectively.  When considering 
the relative information content available in \rave\ (945~pixels in the near 
infrared with $\mathcal{R} \approx 7{,}500$) compared to these literature 
studies that use \project{Hipparcos} parallaxes where possible, and base 
their inferences on spectra with resolving power $\mathcal{R}$ between 
40,000 to 110,000, and S/N ratios exceeding 150, we consider the agreement 
to be very satisfactory.  Indeed, given the metallicity precision available
in the \unrave\ catalog, these results will likely be useful for future 
studies based on exoplanet host star properties (e.g., 
\project{TESS}\footnote{At present, however, there are just $\approx$30 
stars in \rave\ that overlap with the compilations of exoplanet host star 
properties listed at \texttt{exoplanets.org} and \texttt{exoplanets.eu}.}).

@@

There are 142 stars that overlap between \rave\ and the fourth internal
data release of the \ges\ survey. These are a mix of main-sequence, 
sub-giant and red giant branch stars.  About half (67) of the sample 
were acquired with the \acronym{UVES} instrument --- the other with the 
\acronym{GIRAFFE} spectrograph --- and the S/N of the \ges\ spectra peaks 
at $\approx140$~pixel$^{-1}$.  Despite most of these stars having relatively 
low S/N ratios in \rave\ ($\approx 25$~pixel$^{-1}$), there is good agreement
in with \ges\ and the \unrave\ stellar parameters (Figure~\ref{fig:ges-stellar-parameters}).  
The RMS in effective temperature, surface gravity and metallicity is 233~K, 
0.37~dex, and 0.17~dex, respectively.

Based on this comparison, we find no evidence for a systematic offset in
metallicities between stars on the main-sequence and those on the giant branch. 
This is a crucial observation, as the metallicities for stars in our 
main-sequence training set have a principally different source to those on 
the giant branch.  We cannot make these same inferences based on other 
surveys, like \apogee, because (1) \apogee\ stars formed part of the 
training set, and (2) they do not include main-sequence stars.  Even if we 
found good agreement between \epic\ and \apogee\ metallicities, this would 
not be informative, because \apogee\ is the source of metallicity for many 
stars on the giant branch in the \epic\ sample. Therefore, although this is
a qualitative comparison only, it is reassuring that there is no obvious 
systematic difference between the metallicities of main-sequence and giant
branch stars.

The metallicity agreement between this work and \ges\ extends down to low
metallicity, near $[{\rm Fe/H}] \approx -1.5$.  The scatter increases for
the four stars in the overlap sample with [Fe/H] $< -1$, however these
spectra have low S/N ratios, which is reflected in the larger uncertainties
reported for these metallicities.

The fourth internal data release of the \ges\ includes detailed chemical
abundances of up to 45~species ($\approx32$ elements at different ionization
stages).  This provides us with an independent validation for our detailed
abundances on the giant branch.  These comparisons are shown in 
Figure~\ref{fig:ges-abundances}, where markers are colored by the
S/N of the \rave\ spectrum.  The number of stars available in each abundance
comparison varies due to what is available in the \ges\ data release, which
is itself a function of the instrument used, the spectral type, and other
factors.  The absolute bias for individual elements varies from as low as
0.01~dex ([Al/H]) to as high as 0.22~dex ([Si/H]), where we over-estimate 
[Si/H] abundances relative to the \ges\ survey.  The large bias in [Si/H]
is likely a consequence of an offset between [Si/H] abundances in \ges\ and
\apogee, the source of our training set for giant star abundances.  
The RMS deviation in each
label is small for stars with [X/H] $> -0.5$, before increasing at lower
metallicities.  If we consider all stars, the smallest abundance RMS we see
with respect to \ges\ is 0.18~dex for [Ca/H].  The increasing RMS at low 
metallicity is likely a consequence of multiple factors, namely: inaccurate
abundance labels for metal-poor stars (Section~\ref{sec:the-training-set});
only weak, blended lines being available in \rave, which cease to be visible
in hot and/or metal-poor stars; and to a lesser extent, low S/N ratios for 
those particular stars being compared.  Unfortunately not all of these
factors are represented by the quoted errors in each label.  For these reasons, 
although it affects only a small number of stars, we recommend caution when 
using individual abundances for very metal-poor giant stars in our sample.

\subsubsection{Comparison with the \rave\ DR4 calibration sample}
\label{sec:dr4-calibration-sample}

The fourth \rave\ data release made use of a number of high-resolution studies
to verify the accuracy of the reported stellar atmospheric parameters.  These
samples include main-sequence stars, giant stars, with a particular focus to
include metal-poor stars to identify (and correct) any deviations at low
and high metallicities.  We refer the reader to \citet{Kordopatis_2013} for
the full compilation of literature sources.  Although the stellar atmospheric 
parameters in this compilation come from multiple (heterogeneous) sources, we 
find generally good agreement with these works (Figure 
\ref{fig:kordopatis-calibration}).  However, some reservation is warranted 
when evaluating this comparison, as some of the metal-poor stars in this 
calibration sample formed part of our training set.
